---
title: "Mixed models - ACO/CO RI"
author: "HT"
date: "2023-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(ggplot2)
require(ggfortify)
require(MuMIn)
require(gridExtra)
require(car)
require(agricolae)
require(nlme)
require(lme4)
require(lmerTest)
require(emmeans)
```

#########################
**Hybrid Viability data**
#########################

Data input, structure management
```{r}
dat<-read.csv(file.choose()) #choose "Hybrid viability data.csv"
dat$Parental.Replicate<- as.factor(dat$Parental.Replicate)
```

Trait 1 - Development time data 
Step 1: Data visualization
```{r}
#by replicate
ggplot(dat, aes(y = Av.Development.Time, x = Generation, fill= Parental.Replicate)) + 
  geom_boxplot() +
  scale_x_discrete(name= "Cross identity", limits= c("AP", "AF1", "AF2", "CP", "CF1", "CF2"), 
                   labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2")) +
  scale_y_continuous(name= "Average development time (hours)")

#overall population
ggplot(dat, aes(y = Av.Development.Time, x = Generation)) + 
  geom_boxplot() +
  scale_x_discrete(name= "Cross identity", limits= c("AP", "AF1", "AF2", "CP", "CF1", "CF2"), 
                   labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2")) +
  scale_y_continuous(name= "Average development time (hours)")
```

Step 2: Model fitting

Model with cross identity as fixed effect and parental replicate as random effect
```{r}
dev.model<- lmer(Av.Development.Time~Generation + 
                   (1|Parental.Replicate), data=dat)
anova(dev.model) #significant fixed effect 
ranova(dev.model) #no significant random effect
```

Step 3: Model assumptions - visual tests
```{r}
plot(resid(dev.model))
hist(resid(dev.model))
###assumption are met 
```

Step 4: Contrast testing
Using emmeans() to determine pairwise differences 
```{r}
emmeans(dev.model, pairwise~Generation)
```


Step 5: Data summary
```{r}
dev.dat<-dat %>% group_by(Generation) %>% summarize(dev.time.average=mean(Av.Development.Time), dev.time.sd=sd(Av.Development.Time), n=n()); dev.dat
```

Trait 2 - Hatchability data 
Step 1: Data visualization
```{r}
#by replicate
ggplot(dat, aes(y = Hatchability, x = Generation, fill= Parental.Replicate)) + 
  geom_boxplot() +
  scale_x_discrete(name= "Cross identity", limits= c("AP", "AF1", "AF2", "CP", "CF1", "CF2"), 
                   labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2")) +
  scale_y_continuous(name= "Hatchability")

#by population
ggplot(dat, aes(y = Hatchability, x = Generation)) + 
  geom_boxplot() +
  scale_x_discrete(name= "Cross identity", limits= c("AP", "AF1", "AF2", "CP", "CF1", "CF2"), 
                   labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2")) +
  scale_y_continuous(name= "Hatchability")

```

Step 2: Model fitting
A generalized mixed model (binomial), with cross identity as fixed effect and parental replicate as random effect
```{r}
hatch.model<- glmer(Viable.Eggs/Total.Eggs~Generation + (1|Parental.Replicate),
                      weights=Total.Eggs, family=binomial, data=dat)
Anova(hatch.model) #significant fixed effect
#ranova() doesn't work for glmer() so fitting another model without random effect and comparing 
hatch.model.r<- glm(Viable.Eggs/Total.Eggs~Generation, 
                    weights=Total.Eggs, family=binomial, data=dat)
anova(hatch.model, hatch.model.r) #no significant random effect 
```

Step 3: Model assumptions - visual tests
```{r}
plot(resid(hatch.model))
hist(resid(hatch.model))
## assumptions met
```

Step 4: Contrast testing
Using emmeans() to determine pairwise differences 
```{r}
emmeans(hatch.model, pairwise~Generation)
```


Step 5: Data summary
```{r}
dat.hatch<-dat %>% group_by(Generation) %>% summarize(hatch.average=mean(Hatchability), hatch.sd=sd(Hatchability), n=n()); dat.hatch
```

Trait 3 - Larval to adult viability data 
Step 1: Data visualization
```{r}
#by replicate 
ggplot(dat, aes(y = LarvaetoAdultCorrected, x = Generation, fill= Parental.Replicate)) + 
  geom_boxplot() +
  scale_x_discrete(name= "Cross identity", limits= c("AP", "AF1", "AF2", "CP", "CF1", "CF2"), 
                   labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2")) +
  scale_y_continuous(name= "Egg to Adult Viability")

#by population
ggplot(dat, aes(y = LarvaetoAdultCorrected, x = Generation)) + 
  geom_boxplot() +
  scale_x_discrete(name= "Cross identity", limits= c("AP", "AF1", "AF2", "CP", "CF1", "CF2"), 
                   labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2")) +
  scale_y_continuous(name= "Egg to Adult Viability")
```

Step 2: Model fitting
Fitting generalized linear mixed model 
```{r}
viab.model<- glmer(Total.Adults/Viable.Eggs.corrected~Generation+ (1|Parental.Replicate), 
                   weights= Viable.Eggs.corrected, family=binomial, data=dat)
Anova(viab.model) #significant fixed effect

#repeating with lme to determine denominator df
model<-lme(Total.Adults/Viable.Eggs.corrected~Generation, random=~1|Parental.Replicate, data=dat)
anova(model)

#ranova() doesn't work for glmer() so fitting another model without random effect and comparing 
viab.model.r<- glm(Total.Adults/Viable.Eggs.corrected~Generation, data=dat, weights=Viable.Eggs.corrected, family=binomial)
anova(viab.model, viab.model.r) #significant random effect 
```

Step 3: Model assumptions - visual tests
```{r}
plot(resid(viab.model))
hist(resid(viab.model))
## assumptions met
```

Step 4: Contrast testing
Using emmeans() to determine pairwise differences 
```{r}
emmeans(viab.model, pairwise~Generation)
```


Step 5: Data summary
```{r}
dat.viab<-dat %>% group_by(Generation) %>% summarize(viab.average=mean(LarvaetoAdultCorrected), viab.sd=sd(LarvaetoAdultCorrected), n=n());dat.viab
```


#########################
**Hybrid Body Size data**
#########################

Data input, structure management
```{r}
datw<-read.csv(file.choose()) #choose "Hybrid size data.csv"
datw$Parental.Replicate <- as.factor(datw$Parental.Replicate)
```

Step 1: Data visualization
```{r}
#by replicate
ggplot(datw, aes(y = Individual.Weight, x = Generation, fill= Parental.Replicate)) + 
  geom_boxplot() +
  scale_x_discrete(name= "Cross identity", limits= c("AP", "AF1", "AF2", "CP", "CF1", "CF2"), 
                   labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2")) +
  scale_y_continuous(name= "Individual weight") + theme_classic()

#by population
ggplot(datw, aes(y = Individual.Weight, x = Generation)) + 
  geom_boxplot() +
  scale_x_discrete(name= "Cross identity", limits= c("AP", "AF1", "AF2", "CP", "CF1", "CF2"), 
                   labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2")) +
  scale_y_continuous(name= "Individual weight") + theme_classic()
```

Step 2: Model fitting
Fitting linear mixed model 
```{r}
mass.model<- lmer(Individual.Weight~Generation+ (1|Parental.Replicate), data=datw)
anova(mass.model) #singificant fixed effect 
ranova(mass.model) #significant random effect 
```

Step 3: Model assumptions - visual tests
```{r}
plot(resid(mass.model))
hist(resid(mass.model))
## assumptions met
```

Step 4: Contrast testing
Using emmeans() to determine pairwise differences 
```{r}
emmeans(mass.model, pairwise~Generation)
```


Step 5: Data summary
```{r}
dat.mass<-datw %>% group_by(Generation) %>% summarize(weight.average=mean(Individual.Weight), weight.sd=sd(Individual.Weight), n=n());dat.mass
```


###############################
**Postzygotic fertility  data**
###############################

Data input, structure management
```{r}
fert.dat<-read.csv(file.choose()) #choose "PZRF compiled data.csv"
fert.dat<- fert.dat %>% filter(!is.na(Proportion.Red))

#sex based data partition
datf<- fert.dat %>% filter(Sex=="f")
datm<- fert.dat %>% filter(Sex=="m")
```

Trait 1 - Female reproductive fitness data 
Step 1: Data visualization
```{r}
#by replicate 
ggplot(datf) + 
  geom_boxplot(aes(x=Cross.Identity, y=Proportion.Red, fill= Population))+
  scale_x_discrete(name= "Cross identity- female", labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2"))+ 
  scale_y_continuous(name= "Proportion of red eyed offspring")

#by population
ggplot(datf) + 
  geom_boxplot(aes(x=Cross.Identity, y=Proportion.Red))+
  scale_x_discrete(name= "Cross identity- female", labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2"))+ 
  scale_y_continuous(name= "Proportion of red eyed offspring")
```

Step 2: Model fitting
Fitting the generalized linear mixed model, with cross identity as fixed effect and parental replicate as random effect 
```{r}
fem.mod<- glmer(Red/(Red+Brown)~Cross.Identity+ (1|Population), data=datf, weights=(Red+Brown), family=binomial)
Anova(fem.mod) #significant fixed effect 

fem.mod.r<- glm(Red/(Red+Brown)~Cross.Identity, data=datf, weights=(Red+Brown), family=binomial)
anova(fem.mod, fem.mod.r) #no significant random effect 

#repeating with lme to determine denominator df
model.fem<-lme(Red/(Red+Brown)~Cross.Identity, random=~1|Population, data=datf)
anova(model.fem)
```

Step 3: Model assumptions - visual tests
```{r}
plot(resid(fem.mod))
hist(resid(fem.mod)) 
###assumption are met 
```

Step 4: Contrast testing
Using emmeans() to determine pairwise differences 
```{r}
emmeans(fem.mod, pairwise~Cross.Identity)
```


Step 5: Data summary
```{r}
dat.fem<- datf %>% group_by(Cross.Identity) %>% summarize(average= mean(Proportion.Red), sd=sd(Proportion.Red), n=n())
dat.fem
```

Trait 2 - Male reproductive fitness data 
Step 1: Data visualization
```{r}
#by replicate
ggplot(datm) + 
  geom_boxplot(aes(x=Cross.Identity, y=Proportion.Red, fill=Population))+
  scale_x_discrete(name= "Cross identity- male", labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2"))+ 
  scale_y_continuous(name= "Proportion of red eyed offspring")

#by population
ggplot(datm) + 
  geom_boxplot(aes(x=Cross.Identity, y=Proportion.Red))+
  scale_x_discrete(name= "Cross identity- male", labels= c("AfAm", "AfCm F1", "AfCm F2", "CfCm", "CfAm F1", "CfAm F2"))+ 
  scale_y_continuous(name= "Proportion of red eyed offspring")
```

Step 2: Model fitting
Same as females
```{r}
mal.mod<- glmer(Red/(Red+Brown)~Cross.Identity+ (1|Population), data=datm, family= binomial, weights=(Red+Brown))
Anova(mal.mod) #significant effect

mal.mod.r<- glm(Red/(Red+Brown)~Cross.Identity, data=datm, family= binomial, weights=(Red+Brown))
anova(mal.mod, mal.mod.r) #significant effect of replicate 

#repeating with lme to determine denominator df
model.mal<-lme(Red/(Red+Brown)~Cross.Identity, random=~1|Population, data=datm)
anova(model.mal)
```

Step 3: Model assumptions - visual tests
```{r}
plot(resid(mal.mod))
hist(resid(mal.mod)) 
###assumption are met 
```

Step 4: Contrast testing
Using emmeans() to determine pairwise differences 
```{r}
emmeans(mal.mod, pairwise~Cross.Identity)
```


Step 5: Data summary
```{r}
dat.mal<- datm %>% group_by(Cross.Identity) %>% summarize(average= mean(Proportion.Red), sd=sd(Proportion.Red), n=n())
dat.mal
```
